Jane, John … Leslie? A Historical Method for Algorithmic Gender Prediction

### 2014-09-29 article.Rmd
Added plot, plot edits, remapped categories

The path from a dissertation to a book review in the *American Historical Association* is not an automatic one. Not all dissertations are published as books, not all books are submitted to the American Historical Review, and not all book submissions are accepted by the *AHR* editors for review. As of August 2014 the journal's website reminds readers that "the AHR receives over 3,000 books a year; we have the resources to publish at most 1,000 reviews a year (approximately 200 per issue)."^[The website is available at <http://www.indiana.edu/~ahrweb/book_review_guide.html> (retrieved August 25, 2014).] Fewer and fewer female historians make it through each successive step. Establishing a single causal reason for this pattern is all but impossible, and instead a murky concoction of interrelated factors likely contributes to the gap. For one, a 2010 survey by the American Historical Association revealed the female historians dedicated substantially more time to child and elder care than their male colleagues, leaving less time for research. This gender gap is more likely to impact the production of a book than a dissertation, as graduate students are generally younger and less likely to be juggling child-rearing responsibilities while writing than their older peers. More troublingly, women continue to face gender discrimination.

We then used our program to infer the gender of not just book authors reviewed by the *AHR*, but the gender of the reviewers themselves. The story is much the same for reviewers as it is for authors: more than than twice as many men as women appear as reviewers in the journal. But gender inflects book reviews in less direct ways by shaping who writes reviews of which book authors. About three times as many men write reviews of male-authored books as do women. In the case of female-authored books the ratio of female-to-male reviewers is much closer to 50/50. In short, women are much more likely to write reviews of other women. And while men still write the majority of reviews of female-authored books, they tend to gravitate more towards male authors – who are, of course, already over-represented in the *AHR*.

```{r}
ahr_predicted %>%
  filter(gender_author == "female") %>%
  select(year, gender_reviewer) %>%  
  group_by(year, gender_reviewer) %>% 
  summarize(n = n()) %>% 
  spread(gender_reviewer, n, fill = 0) %>%
  mutate(female_proportion = female / (female + male + unknown),
         male_proportion = male / (female + male + unknown),
         unknown_proportion = unknown / (female + male + unknown)) %>%  
  select(female_proportion, male_proportion, unknown_proportion, year) %>%
  gather(gender, proportion, -year) %>%
  ggplot(aes(x = year, y = proportion, fill = gender)) + 
  geom_bar(binwidth = 1, stat = "identity") +
  ggtitle("Gender of reviewers of female book authors in AHR") +
  xlab(NULL) + ylab(NULL) +
  theme(legend.position = "bottom") +
  geom_hline(yintercept = 0.5) +
  scale_fill_brewer(palette = article_palette, 
                    labels = c("Female", "Male", "Unknown")) +
  article_theme +
  short_scale
```

The American Historical Association also categorizes their reviews under different historical sub-fields. Although the journal’s taxonomy has changed quite a bit over time, it did allow us to refine our analysis. Were there more reviews of female authors in certain sub-fields rather than others, and has that changed over the past forty years? Caribbean/Latin American history has had something approaching equal representation for the past decade-and-a-half. In both African history and Ancient/Medieval history female historians made some quite dramatic gains during the late-nineties and aughts. The most uneven parties, however, are also the two subject categories that publish the most book reviews in the *AHR*: Modern/Early Modern Europe and the United States/Canada. Both sub-fields have made progress over the four decades but still hover at around two-thirds male. Gender representation is not evenly distributed across the profession.

@@

ahr_predicted %>%
  left_join(categories, by = "category") %>%
  filter(gender_author == "female",
         !our_category %in% c(NA, 
                              "Latin and South America, Carribean",
                              "Middle East")) %>%
  select(year, gender_reviewer, our_category) %>% 
  group_by(year, gender_reviewer, our_category) %>%  
  summarize(n = n()) %>%
  spread(gender_reviewer, n, fill = 0) %>% 
  mutate(female_proportion = female / (female + male + unknown),
         male_proportion = male / (female + male + unknown),
         unknown_proportion = unknown / (female + male + unknown)) %>%  
  select(female_proportion, male_proportion, unknown_proportion,
         year, our_category) %>% 
  gather(gender, proportion, -year, -our_category) %>%
ggplot(aes(x = year, y = proportion, fill = gender)) + 
  geom_bar(binwidth = 1, stat = "identity") +
  facet_wrap(~ our_category) +
  xlim(1970, 2012) +
  ggtitle("Gender of reviewers of female AHR authors by category, 1970-2012") +
  xlab(NULL) + ylab(NULL) +
  theme(legend.position = "bottom") +
  geom_hline(yintercept = 0.5) +
  scale_fill_brewer(palette = article_palette,
                    labels = c("Female", "Male", "Unknown"),
                    name = "Gender") +
  article_theme +
  theme(strip.background = element_rect(color = "white"))
```

>>>>

The path from a dissertation to a book review in the *American Historical Association* is not automatic. Not all dissertations are published as books, not all books are submitted to the *American Historical Review*, and not all book submissions are accepted by the *AHR* editors for review. As of September 2014 the journal's website reminds readers that "the AHR receives over 3,000 books a year; we have the resources to publish at most 1,000 reviews a year (approximately 200 per issue)." [American Historical Review n.d.] Fewer and fewer female historians make it through each successive step. Establishing a single causal reason for this pattern is all but impossible, and instead a murky concoction of interrelated factors likely contributes to the gap. For one, a 2010 survey by the American Historical Association revealed the female historians dedicated substantially more time to child and elder care than their male colleagues, leaving less time for research. This gender gap is more likely to impact the production of a book than a dissertation, as graduate students are generally younger and less likely to be juggling child-rearing responsibilities while writing than their older peers. More troublingly, women continue to face gender discrimination in the professional sphere.

We  used our program to infer the gender of not just book authors reviewed by the *AHR*, but the gender of the reviewers themselves. The story is much the same for reviewers as it is for authors: more than than twice as many men as women appear as reviewers in the journal. But gender inflects book reviews in less direct ways by shaping who writes reviews of which book authors. About three times as many men write reviews of male-authored books as do women. In the case of female-authored books the ratio of female-to-male reviewers is much closer to 50/50. In short, women are much more likely to write reviews of other women. And while men still write the majority of reviews of female-authored books, they tend to gravitate more towards male authors – who are, of course, already over-represented in the *AHR*.

```{r}
ahr_predicted %>%
  filter(gender_author == "male") %>%
  select(year, gender_reviewer) %>%  
  group_by(year, gender_reviewer) %>% 
  summarize(n = n()) %>% 
  spread(gender_reviewer, n, fill = 0) %>%
  mutate(female_proportion = female / (female + male + unknown),
         male_proportion = male / (female + male + unknown),
         unknown_proportion = unknown / (female + male + unknown)) %>%  
  select(female_proportion, male_proportion, unknown_proportion, year) %>%
  gather(gender, proportion, -year) %>%
  ggplot(aes(x = year, y = proportion, fill = gender)) + 
  geom_bar(binwidth = 1, stat = "identity") +
  ggtitle("Gender of reviewers of MALE book authors in AHR") +
  xlab(NULL) + ylab(NULL) +
  theme(legend.position = "bottom") +
  geom_hline(yintercept = 0.5) +
  scale_fill_brewer(palette = article_palette, 
                    labels = c("Female", "Male", "Unknown")) +
  article_theme +
  short_scale
```

```{r}
ahr_predicted %>%
  filter(gender_author == "female") %>%
  select(year, gender_reviewer) %>%  
  group_by(year, gender_reviewer) %>% 
  summarize(n = n()) %>% 
  spread(gender_reviewer, n, fill = 0) %>%
  mutate(female_proportion = female / (female + male + unknown),
         male_proportion = male / (female + male + unknown),
         unknown_proportion = unknown / (female + male + unknown)) %>%  
  select(female_proportion, male_proportion, unknown_proportion, year) %>%
  gather(gender, proportion, -year) %>%
  ggplot(aes(x = year, y = proportion, fill = gender)) + 
  geom_bar(binwidth = 1, stat = "identity") +
  ggtitle("Gender of reviewers of FEMALE book authors in AHR") +
  xlab(NULL) + ylab(NULL) +
  theme(legend.position = "bottom") +
  geom_hline(yintercept = 0.5) +
  scale_fill_brewer(palette = article_palette, 
                    labels = c("Female", "Male", "Unknown")) +
  article_theme +
  short_scale
```

The *AHR* also categorizes its reviews under different historical sub-fields. Although the journal’s taxonomy has changed quite a bit over time, it did allow us to refine our analysis. Were there more reviews of female authors in certain sub-fields rather than others, and has that changed over the past forty years? Caribbean/Latin American history has had something approaching equal representation for the past decade-and-a-half. In both African history and Ancient/Medieval history female historians made some quite dramatic gains during the late-nineties and aughts. The most uneven parties, however, are also the two subject categories that publish the most book reviews in the *AHR*: Modern/Early Modern Europe and the United States/Canada. Both sub-fields have made progress over the four decades but still hover at around two-thirds male. Gender representation is not evenly distributed across the profession.

@@

ahr_predicted %>%
  left_join(categories, by = "category") %>%
  filter(!our_category %in% c(NA, 
                              "Articles",
                              "Latin and South America, Carribean",
                              "Middle East")) %>%
  select(year, gender_author, our_category) %>% 
  group_by(year, gender_author, our_category) %>%  
  summarize(n = n()) %>%
  spread(gender_author, n, fill = 0) %>% 
  mutate(female_proportion = female / (female + male + unknown),
         male_proportion = male / (female + male + unknown),
         unknown_proportion = unknown / (female + male + unknown)) %>%  
  select(female_proportion, male_proportion, unknown_proportion,
         year, our_category) %>% 
  gather(gender, proportion, -year, -our_category) %>%
ggplot(aes(x = year, y = proportion, fill = gender)) + 
  geom_bar(binwidth = 1, stat = "identity") +
  facet_wrap(~ our_category) +
  xlim(1970, 2012) +
  ggtitle("Gender of AHR authors by book category, 1970-2012") +
  xlab(NULL) + ylab(NULL) +
  theme(legend.position = "bottom") +
  geom_hline(yintercept = 0.5) +
  scale_fill_brewer(palette = article_palette,
                    labels = c("Female", "Male", "Unknown"),
                    name = "Gender") +
  article_theme +
  theme(strip.background = element_rect(color = "white"))
```
